---
title: "README"
author: "Kamil Bojanczyk"
date: "May 25, 2015"
output: md_document
keep_md: TRUE
---

# Code Book of Variables and Script Explanation

This document describes the codebook of variables from the UCLA Samsung Galaxy S [data set.](http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones) The data has been cleaned and cut down into a final form, which is available in the [output.txt file.](https://github.com/kamilien1/get-clean-final/blob/master/output.txt)

## Script Explanation  

The `run_analysis.R` script performs five basic functions:  
1. Merges training and test data sets  
2. Extracts mean and standard deviation measurements from the data sets  
3. Provides descriptive names for activities 1 to 6  
4. Labels the measurements with appropriate, descriptive names  
5. Writes an output file of average values of sd and mean measurements, by **test subject** and by **activity type**  


Top level, the script does the steps in the following order:  
1. Load and process test data  
2. Load and process train data  
3. Merge the data sets
4. Take the mean of the values of interest, by test subject
5. Write the file into output_ppl.txt

## Step 1: Look at the Raw Data

*Note: The Inertial Signals data folder was not needed for this assignment, and thus is ignored.*


```{r echo=TRUE, cache=TRUE}
test = "../data/test"
list.files("../data")
list.files(test)

```

Looking at the files, we see **generic** and **specific** files. 


Generic files that we care about are:  
- `activity_labels.txt`: the 6 activities with text labels  
- `features.txt`: the 561 features calucated from XYZ data


The three files specific to test/train data, for test data are:  
- `subject_test.txt`: Subject 1 through 30, showing what subject was tested  
- `X-test.txt`: Specific data points for the 561 features calcuations performed  
- `Y-test.txt`: The label of the 6 tests




## Step 2: Process the Test Data

Let's get the test data and process it.

```{r echo=TRUE, cache=TRUE}
# get the subject data 
sub_t <- "../data/test/subject_test.txt"
sub_tx <- "../data/test/X_test.txt"
sub_ty <- "../data/test/Y_test.txt"

```

### Process the People Data

Read in the subject test people, 1 through 30. 

```{r echo=TRUE, cache=TRUE}
# subject is the subject (1:30) who performed the activity
st <- read.table(sub_t,header=F,stringsAsFactors=F)
# change the names to person
names(st) <- "person"
st$person <- factor(st$person)
dim(st)

```

### Process the 561 Feature Data

```{r echo=TRUE, cache=TRUE}

# analyzed data on activities/features performed
stx <- read.table(sub_tx,header=F,stringsAsFactors=F)

```


### Process the Activity Data

``` {r echo=TRUE, cache=TRUE}

# labels of each type of activity (1:6)
sty <- read.table(sub_ty,header=F,stringsAsFactors=T)
names(sty) <- "activity"
sty$activity <- factor(sty$activity)

# names of the 1:6 activities
alab <- "../data/activity_labels.txt"
albl <- read.table(alab,header=F,stringsAsFactors=T)
albl$V1 <- factor(albl$V1)

```


### Merge Data

Merging is a multi-step process, shown below. We merge the people to their activity first, and then we further merge `sty_labeled_ppl` with the 561 features. 

``` {r echo=TRUE, cache=TRUE}

sty_labeled_ppl <- cbind(factor(sty$activity,labels=albl$V2),st)
# TODO: remove activity column? 
names(sty_labeled_ppl) <- c("activity","person")
head(sty_labeled_ppl)
# fully merged
test_data_ppl <- cbind(sty_labeled_ppl,stx)
dim(test_data_ppl)

# load the features and label the columns 
features <- "../data/features.txt"
fts <- read.table(features,header=F, stringsAsFactors=F)

# test_data_ppl[1:2] is activity # and person #
# test_data_ppl[3:563] is the 561 set of features calculated
# to make life easier, i'll make these column names be their names
names(test_data_ppl)[3:563] <- fts$V2
# sanity check output
names(test_data_ppl)[c(1:3,560:563)]

```


### Use Grep() to Find Mean and SD Data

In the final step, we use `grep()` to find mean and standard dev data.

```{r echo=TRUE, cache=TRUE}

new_names_ppl <- names(test_data_ppl)[grep("mean\\(|std\\(|activity|person",names(test_data_ppl))]
# subset the test data with the new, grepped colnames
test_data_complete_ppl <- test_data_ppl[,new_names_ppl]

# sanity test
names(test_data_complete_ppl)

```




## Step 3: Process the Train Data

This is the same as **step 2**, except for train data. Only the final size of the data frame is displayed. 

```{r echo=FALSE, cache=TRUE}

########### same exact thing for the train data!
# file reads
sub_tr <- "../data/train/subject_train.txt"
sub_trx <- "../data/train/X_train.txt"
sub_try <- "../data/train/Y_train.txt"

str <- read.table(sub_tr,header=F,stringsAsFactors=F)
names(str) <- "person"
str$person <- factor(str$person)

strx <- read.table(sub_trx,header=F,stringsAsFactors=F)

# labels of each type of activity (1:6)
stry <- read.table(sub_try,header=F,stringsAsFactors=T)
names(stry) <-"activity"
stry$activity <- factor(stry$activity)

# stry and albl must match up

stry_labeled_ppl <- cbind(factor(stry$activity,labels=albl$V2),str)
names(stry_labeled_ppl) <- c("activity","person")

train_data_ppl <- cbind(stry_labeled_ppl,strx)
dim(train_data_ppl)

# apply fts to names
names(train_data_ppl)[-(1:2)] <- fts$V2

# now train data has properly labeled columns and rows
train_data_complete_ppl <- train_data_ppl[,new_names_ppl]
##########



```


## Step 3: Merge Test and Train, Get Output Data

Now that we have cleaned up our data, let's merge it one more time into `complete_data_set_ppl`. Then we'll use aggregate() to aggregate all data by `activity` and by `person`, and use the mean() function to take the average of each feature **by** activtity **and** person. The output will be 180 rows (30 people, 6 activities) by 68 columns (activity, person, and 66 standard dev or mean features).

``` {r echo=TRUE, cache=TRUE}


complete_data_set_ppl <- rbind(test_data_complete_ppl,train_data_complete_ppl)
## now data is "cleaned up" 

# average all columns
# by activity and act_name, which are the same
ans_ppl <- aggregate(.~activity+person,data=complete_data_set_ppl,FUN=mean)
dim(ans_ppl)
ans_ppl[1:3,1:4]

# write table
# http://www.statmethods.net/input/exportingdata.html
write.table(ans_ppl, "output_ppl.txt",sep="\t",row.names=FALSE)


```

